% Sample LaTeX file for creating a paper in the Morgan Kaufmannn two
% column, 8 1/2 by 11 inch proceedings format.

\documentclass[]{article}
\usepackage{proceed2e}
\usepackage{graphicx}
\usepackage{amsmath}

% Set the typeface to Times Roman
\usepackage{times}

\usepackage{hyperref}
\usepackage{url}

\input{preamble.tex}

\title{Scalable Recommendation with Hierarchical Poisson
  Factorization}
\author{ {\bf Prem Gopalan} \\
Department of Computer Science \\
Princeton University \\
Princeton, NJ \\
\And
{\bf Jake M. Hofman} \\
Microsoft Research \\
641 Sixth Avenue, Floor 7 \\
New York, NY \\
\And
{\bf David M. Blei} \\
Departments of Statistics \& Computer Science \\
Columbia University \\
New York, NY \\
}

% The author names and affiliations should appear only in the accepted paper.
%
%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
%alternate address.}} \\
%Computer Science Dept. \\
%Cranberry University\\
%Pittsburgh, PA 15213 \\
%\And
%{\bf Coauthor}  \\
%Affiliation          \\
%Address \\
%\And
%{\bf Coauthor}   \\
%Affiliation \\
%Address    \\
%(if needed)\\
%}

\begin{document}

\maketitle

\begin{abstract}
We develop hierarchical Poisson matrix factorization (HPF), a novel
method for providing users with high quality recommendations based on
implicit feedback, such as views, clicks, or purchases.  In contrast
to existing recommendation models, HPF has a number of desirable
properties.  First, we show that HPF more accurately captures the
long-tailed user activity found in most consumption data by explicitly
considering the fact that users have finite attention budgets.  This
leads to better estimates of users' latent preferences, and therefore
superior recommendations, compared to competing methods.  Second, HPF
learns these latent factors by only explicitly considering positive
examples, eliminating the often costly step of generating artificial
negative examples when fitting to implicit data.  Third, HPF is more
than just one method---it is the simplest in a class of probabilistic
models with these properties, and can easily be extended to include
more complex structure and assumptions.  We develop a variational
algorithm for approximate posterior inference for HPF that scales up
to large data sets, and we demonstrate its performance on a wide
variety of real-world recommendation problems---users rating movies,
listening to songs, reading scientific papers, and reading news
articles.  
%% Our study reveals that hierarchical Poisson factorization
%% outperforms previous methods, including nonnegative matrix
%% factorization, topic models, and probabilistic matrix factorization
%% techniques.
\end{abstract}

\input{intro.tex}
\input{model.tex}
\input{inference.tex}
\input{eval.tex}
\input{related.tex}
\input{discussion.tex}

\subsubsection*{Acknowledgements}
We thank Ulrich Paquet, Laurent Charlin, Francisco J. R. Ruiz, Rajesh
Ranganath, Matthew D. Hoffman and the anonymous reviewers for comments
and discussions. DMB is supported by NSF CAREER NSF IIS-0745520, NSF
BIGDATA NSF IIS-1247664, ONR N00014-11-1-0651, and DARPA
FA8750-14-2-0009.

\small{
  \bibliographystyle{abbrv}
  \bibliography{bib}
}

\newpage
\appendix
\input{appendix.tex}

\end{document}
